Model to study co-evolution of trade an culture

### 2015-04-06 article.tex
small fixes

The first aspect of our model is focused on the dynamics of cultural change. They comprise the collection of processes that promotes or inhibits the spread of information by social interaction within a population \cite[3]{boyd_origin_2005}. An increasing number of social scientists are using an evolutionary framework to model these dynamics, thus fostering the development of transdisciplinary efforts designed to understand cultural evolution \cite{henrich_evolution_2003}. Archaeology has a rich tradition of evolutionary studies \cite{lycett_cultural_2015}, exploring concepts such as gene-culture coevolution \cite{burger_absence_2007}, diffusion processes \cite{fort_synthesis_2012}, phylogenetic analysis of material culture \cite{obrien_cladistics_2001} or detection of rates of change in the archaeological record \cite{premo_cultural_2014}.

A large percentage of these studies focus on the biases that affect the transmission of cultural traits, including the relevance of intrinsic traits and frequency dependence (e.g. conformism and anti-conformism). 
These analysis are often based on the theory of neutral evolution \cite{neiman_stylistic_1995}, which explains variation in the distribution of traits as a function of their frequency distribution in the population. Thus, a trait is assumed to lack any selective bias and its transmission will depend on its popularity amongst the rest of individuals. The comparison of a given frequency distribution found in data against this unbiased transmission model is traditionally used as a null-hypothesis testing for selection. There exist a large number of works using this approach as the starting point to analyse patterns of cultural change in past societies \cite{lipo_neutralitystyle_2001,shennan_ceramic_2001,steele_ceramic_2010,kandler_nonequilibrium_2013,porcic_exploring_2014,crema_approximate_2014}.

Unbiased transmission generates a clear pattern of power law distributions that can be replicated with a simple random copying transmission mechanism \cite{bentley_random_2004}. Within this framework, an individual will copy the traits of a randomly chosen individual with a given probability. This copy can potentially introduce some errors in the acquired trait, which account for innovation processes. The individual will in turn continue to spread these cultural traits which will be further adopted by other individuals. This basic model can be enriched by several additional processes both in the innovation \cite{schillinger_copying_2014,sole_evolutionary_2013,ziman_technological_2003} and the transmission \cite{heyes_social_1994,henrich_evolution_2003}.

@@

Previous historical and archaeological studies to model trade in the ancient Mediterranean societies have focus on descriptive approaches ~\cite{hopkins_taxes_1980,temin_market_2001,terpstra_trade_2011,temin_economy_2006,wilson_approaches_2009,scheidel_model_2007,kessler_organization_2007}. 
These works have conclude that the Roman economy was a type of market economy where a variety of small tightly connected markets had loose connections between them~\cite{temin_market_2001,temin_economy_2006}. 
Based on these conclusions, further works have attempt to define more exactly the characteristics of this market such, e.g. level of intervention of the state and its impact, or the relation-ship between local and long-distance trade~\cite{wilson_approaches_2009}.

@@

The model proposed is an ABM based on agents and goods. The interactions between agents and goods are modelled thanks to the four processes (production, cultural transmission, innovation and trade) which will be explained in further details later. The flexibility of the models comes from the possibility to implement these processes differently depending on the hypothesis tested. However, since all experiments are performed with the same four processes, the effect of the various implementations of one process can be studied easily. Moreover, the operating of the model revolves around the value that agents associate to products. Depending on the question studied, the value can reflect the interest of the agent for the good or it can be the price at which the agent evaluate this good. Since this vector is common to all experiments performed with this model, it is extremely useful for the comparisons of results obtained with different implementation of the key processes.

@@

More formally the model is composed of a population $Pop$ of $m$ agents, each defined by 2 vectors of size $n$. The first corresponds to the quantity of each good that the agent $i$ possesses: 
$$\forall i \in Pop, \quad Q^i = (q^i_1,\cdots,q^i_n) $$

@@

This scheme allow us to have a model achieving the two main goals listed in section~\ref{sec:intro}. The first goal is to be able to implement and compare various cultural transmission mechanisms as well as trade mechanisms within the same framework. We argue that this is possible in our framework by modifying only these four key processes i.e. production, trade, imitation and innovation. Moreover, thanks to the use of a unique $V^i$ vector, it is possible to compare the outcomes of different combinations of production, imitation and innovation mechanisms. The second goal is to produce a model composed of a low number of parameter. Our model is composed of only four processes and two vectors which is significantly less complex than previous models.

@@

\begin{figure}[htp]
	\begin{center}
		\includegraphics[width=7cm]{img/fitness.png}
	\end{center}
	\caption{The fitness depending on price and needs}
	\label{fig:fit}
\end{figure}
The figure~\ref{fig:fit} depicts the score of an agent, for different need value of one good, with regards to the quantity of the good possessed by the agent. We can see that even if the need for a good is small (for instance when $need(x) = .01$) an agent can have a high score as soon as it manages to get a quantity close to the need. This insures that an agent needs to have for each of the good a quantity equal to the need and that no agent could reach a better score just by getting the right quantity for a good with a high need.

@@

Once de quantities are defined, the agents declare that the trade is possible if :
\begin{align}
q_g^o >= w_g^o \label{eq:constraintFirst}\\
q_g^r <= w_g^o \\
q_k^r >= w_k^o \label{eq:constraintLastInv}\\
w_g^o>=(q_g^r+w_g^r)\label{eq:constraintFirstWill} \\
w_k^o<=w_k^r \\
w_g^o<=w_g^r
\label{eq:constraintLast}
\end{align}

The conditions \ref{eq:constraintFirst}-\ref{eq:constraintLastInv} insure that both agents have enough goods in their inventory to realise the trade and the conditions \ref{eq:constraintFirstWill}-\ref{eq:constraintLast} insure that the quantities of each good fit the will of both agents.

@@

Funding for this work was provided by the ERC Advanced Grant EPNet (\todo{look for code}) and the SimulPast Consolider Ingenio project (CSD2010-00034) of the former Ministry for Science and Innovation of the Spanish Government.

>>>>

The first aspect of our model is focused on the dynamics of cultural change. They comprise the collection of processes that promotes or inhibits the spread of information by social interaction within a population~\cite[3]{boyd_origin_2005}. An increasing number of social scientists are using an evolutionary framework to model these dynamics, thus fostering the development of transdisciplinary efforts designed to understand cultural evolution~\cite{henrich_evolution_2003}. Archaeology has a rich tradition of evolutionary studies~\cite{lycett_cultural_2015}, exploring concepts such as gene-culture coevolution~\cite{burger_absence_2007}, diffusion processes~\cite{fort_synthesis_2012}, phylogenetic analysis of material culture~\cite{obrien_cladistics_2001} or detection of rates of change in the archaeological record~\cite{premo_cultural_2014}.

A large percentage of these studies focus on the biases that affect the transmission of cultural traits, including the relevance of intrinsic traits and frequency dependence (e.g. conformism and anti-conformism). 
These analysis are often based on the theory of neutral evolution~\cite{neiman_stylistic_1995}, which explains variation in the distribution of traits as a function of their frequency distribution in the population. Thus, a trait is assumed to lack any selective bias and its transmission will depend on its popularity amongst the rest of individuals. The comparison of a given frequency distribution found in data against this unbiased transmission model is traditionally used as a null-hypothesis testing for selection. There exist a large number of works using this approach as the starting point to analyse patterns of cultural change in past societies~\cite{lipo_neutralitystyle_2001,shennan_ceramic_2001,steele_ceramic_2010,kandler_nonequilibrium_2013,porcic_exploring_2014,crema_approximate_2014}.

Unbiased transmission generates a clear pattern of power law distributions that can be replicated with a simple random copying transmission mechanism~\cite{bentley_random_2004}. Within this framework, an individual will copy the traits of a randomly chosen individual with a given probability. This copy can potentially introduce some errors in the acquired trait, which account for innovation processes. The individual will in turn continue to spread these cultural traits which will be further adopted by other individuals. This basic model can be enriched by several additional processes both in the innovation~\cite{schillinger_copying_2014,sole_evolutionary_2013,ziman_technological_2003} and the transmissio~\cite{heyes_social_1994,henrich_evolution_2003}.

@@

Previous historical and archaeological studies to model trade in the ancient Mediterranean societies have focus on descriptive approaches~\cite{hopkins_taxes_1980,temin_market_2001,terpstra_trade_2011,temin_economy_2006,wilson_approaches_2009,scheidel_model_2007,kessler_organization_2007}. 
These works have conclude that the Roman economy was a type of market economy where a variety of small tightly connected markets had loose connections between them~\cite{temin_market_2001,temin_economy_2006}. 
Based on these conclusions, further works have attempt to define more exactly the characteristics of this market such, e.g. level of intervention of the state and its impact, or the relation-ship between local and long-distance trade~\cite{wilson_approaches_2009}.

@@

To explore this coevolution between trade and cultural change we have developed an ABM where the different agents trade goods with variable prices.

\xrc{we can totally remove this paragraph and move the entire next subsection here:\\
These dynamics are modelled after processes (production, cultural transmission, innovation and trade) which will be explained in further details later. The flexibility of the models comes from the possibility to implement these processes differently depending on the hypothesis tested. However, since all experiments are performed with the same four processes, the effect of the various implementations of one process can be studied easily. Moreover, the operating of the model revolves around the value that agents associate to products. Depending on the question studied, the value can reflect the interest of the agent for the good or it can be the price at which the agent evaluate this good. Since this vector is common to all experiments performed with this model, it is extremely useful for the comparisons of results obtained with different implementation of the key processes.}

@@

The model is composed of a population $Pop$ of $m$ agents, each defined by 2 vectors of size $n$. The first corresponds to the quantity of each good that the agent $i$ possesses: 
$$\forall i \in Pop, \quad Q^i = (q^i_1,\cdots,q^i_n) $$

@@

This scheme achieves the two main goals listed in section~\ref{sec:intro}. The first goal is to be able to implement and compare various cultural transmission mechanisms as well as trade mechanisms within the same framework. We argue that this is possible in our framework by modifying only these four key processes i.e. production, trade, imitation and innovation. Moreover, thanks to the use of a unique $V^i$ vector, it is possible to compare the outcomes of different combinations of production, imitation and innovation mechanisms. The second goal is to produce a model composed of a low number of parameter. Our model is composed of only four processes and two vectors which is significantly less complex than previous models.

\xrc{I would remove the previous paragraph, because it is implied that we want few parameters and compare mechanisms}

@@

\begin{figure}[htp]
	\begin{center}
		\includegraphics[width=7cm]{img/fitness.png}
	\end{center}
	\caption{The fitness depending on price and needs}
	\label{fig:fit}
\end{figure}
The figure~\ref{fig:fit} depicts the score of an agent, for different need value of one good, with regards to the quantity of the good possessed by the agent. We can see that even if the need for a good is small (for instance when $need(x) = 0.01$) an agent can have a high score as soon as it manages to get a quantity close to the need. This insures that an agent needs to have for each of the good a quantity equal to the need and that no agent could reach a better score just by getting the right quantity for a good with a high need.

@@

Once the quantities are defined, the agents declare that the trade is possible if :
\begin{align}
q_g^o >= w_g^o \label{eq:constraintFirst}\\
q_g^r <= w_g^o \\
q_k^r >= w_k^o \label{eq:constraintLastInv}\\
w_g^o>=(q_g^r+w_g^r)\label{eq:constraintFirstWill} \\
w_k^o<=w_k^r \\
w_g^o<=w_g^r
\label{eq:constraintLast}
\end{align}

The conditions \ref{eq:constraintFirst}-\ref{eq:constraintLastInv} insure that both agents have enough goods in their inventory to realise the trade while the conditions \ref{eq:constraintFirstWill}-\ref{eq:constraintLast} insure that the quantities of each good fit the will of both agents.

@@

Funding for this work was provided by the ERC Advanced Grant EPNet (340828) and the SimulPast Consolider Ingenio project (CSD2010-00034) of the former Ministry for Science and Innovation of the Spanish Government.
